MISUSES AND INACCURACIES OF PAGEL'S LAMBDA ML ESTIMATION METHODS

### 2020-04-25 Manuscript/Manuscript-2.tex
vsn 2

Evolutionary biologists commonly seek to describe the relative strength
of phylogenetic signal in phenotypic traits, to determine the extent to
which shared evolutionary history has influenced trait covariation among
taxa. This is often accomplished by interpreting empirical estimates of
\(\lambda\); with smaller values signifying `weak' phylogenetic signal,
while larger values are interpreted as `strong' phylogenetic signal
(e.g., De Meester et al. 2019; Pintanel et al. 2019; Su et al. 2019).
Other approaches for interpreting \(\lambda\) are more statistical,
through the use of confidence intervals (Vandelook et al. 2019) or
likelihood ratio tests that compare the observed model fit to that
obtained when \(\lambda=0\) or \(\lambda=1\) (Freckleton et al. 2002;
Cooper et al. 2010; Bose et al. 2019). Likewise, qualitative comparisons
of \(\lambda\) across multiple phenotypic traits have also been used to
infer whether the strength of phylogenetic signal is greater in one
trait as compared to another (e.g., Liu et al. 2019; Bai et al. 2019).
Indeed, it seems intuitive to interpret the strength of phylogenetic
signal in this manner, as \(\lambda\) is a parameter on a bounded scale
(\(0\to1\)) for which interpretation of its extremal points are
understood (\(\lambda=0\) represents no phylogenetic signal, while
\(\lambda=1\) is phylogenetic signal as expected under Brownian motion).
However, equating values of \(\lambda\) directly to the strength of
phylogenetic signal presumes two important statistical properties that
have not been fully explored. \hfill\break

First, it presumes that values of \(\lambda\) can be precisely
estimated, as biological inferences regarding the strength of
phylogenetic signal depend on high accuracy in its estimation.
Therefore, understanding the precision in estimating \(\lambda\) is
paramount. One study (Boettiger et al. 2012) found that estimates of
Pagel's \(\lambda\) displayed less variation (i.e., greater precision)
when data were simulated on a large phylogeny (\(N=281\)) as compared to
a small one (\(N=13\)). From this observation it was concluded that
insufficient data (i.e., the number of species) was the underlying cause
of the increased variation across parameter estimates (Boettiger et al.
2012). Indeed, such a pattern is common with statistical estimators, as
summary statistics and parameters are often more precise at greater
sample sizes (Cohen 1988). However, this conclusion also assumes that
the precision of \(\lambda\) remains constant across its range
(\(\lambda = 0 \to 1\)); an assumption that to date, has not been
verified. Thus, despite widespread use of Pagel's (1999) \(\lambda\) in
macroevolutionary studies, at present, we still lack a general
understanding of the precision with which \(\lambda\) can estimate
levels of phylogenetic signal in phenotypic datasets. \hfill\break

@@

\emph{Results.} We found that the precision of \(\lambda_{est}\) varied
widely across simulation conditions. Predictably, precision improved as
the number of species increased (Figure 1). This confirmed earlier
findings of Boettiger et al.~(2012), and adhered to parametric
statistical theory. However, in many cases the set of \(\lambda_{est}\)
spanned nearly the entire range of possible values (e.g., \(n=32\);
\(\lambda_{in}=0.5\): \(\lambda_{est}= 0.0\to 0.985\)), revealing that
estimates of \(\lambda\) were not a reliable indicator of input
phylogenetic signal. Importantly, the precision of \(\lambda_{est}\) was
not uniform across all levels of phylogenetic signal, with the worst
precision at intermediate levels of signal (\(\lambda_{in}\approx0.5\)),
and improved precision as input levels approached the extremes of its
range (i.e., \(\lambda_{in}\rightarrow0\) \&
\(\lambda_{in}\rightarrow1\)). Thus, estimates of \(\lambda\) were least
reflective of the true input signal at intermediate values.
Additionally, even at large levels of species richness, we found that
the range of \(\lambda_{est}\) still encompassed a substantial portion
of possible values (e.g., \(n=512\); \(\lambda_{in}=0.5\):
\(\lambda_{est} = 0.32\to 0.68\)). Likewise, the same \(\lambda_{est}\)
could be obtained from datasets containing vastly different input levels
of phylogenetic signal (e.g., \(n=512\); \(\lambda_{est} = 0.5\);
\(\lambda_{in} = 0.25 \to 0.65\)). Results were similar when \(\lambda\)
was co-estimated with regression parameters in PGLS regression (Figure
2). Here, regression parameters (\(\beta\)) were accurately estimated,
confirming earlier findings of Revell 2010 (2010) (see Supporting
Information). However, estimates of phylogenetic signal were not, and
the spread of \(\lambda_{est}\) was even broader than that observed when
\(\lambda\) was estimated for only the dependent variable. Taken
together, these findings reveal that \(\lambda_{est}\) does not
precisely characterize observed levels of phylogenetic signal in
phenotypic datasets, and that biological interpretations of the strength
of phylogenetic signal based on \(\lambda\) may be highly inaccurate.
\hfill\break

@@

@@

Adams and Collyer (2019b) recently suggested that the strength of
phylogenetic signal could be represented an effect size, obtained from
\(Kappa\) and its empirical sampling distribution from permutation. Here
we formalize that suggestion, and find an effect size as:

@@

Similarly, an effect size based on \(\lambda\) could be envisioned as:

\begin{align}
    Z_{\lambda}=\frac{\lambda_{obs}-0}{\hat\sigma_\lambda}
\end{align}

Note that under the null hypothesis, \(E(\lambda)=0\), a no phylogenetic
signal is expected under this condition (Freckleton et al. 2002).
\hfill\break

To evaluate the utility of \(Z_K\) and \(Z_\lambda\) we calculated both
effect sizes for the simulated datasets generated above, and summarized
the precision of each using its variance (\(\sigma^2_{Z_K}\) and
\(\sigma^2_{Z_\lambda}\)). Results are found in Figure 3. Here two
things are evident. First, estimates of \(Z_K\) track the input
phylogenetic signal in a more linear fashion than to estimates of
\(Z_\lambda\). Second, the precision of \(Z_K\) is considerably more
stable, as coefficients of variation for the set of \(\sigma^2_{Z_K}\)
across input levels of phylogenetic signal for the same sample size was
an order of magnitude smaller than that observed for
\(\sigma^2_{Z_\lambda}\) (Figure 3). This implies that estimates of the
strength of phylogenetic signal were more reliable and robust when using
\(Z_K\) as compared with \(Z_\lambda\). \hfill\break

{[}insert Figure 3 here{]}

@@

\hypertarget{literature-survey-where-does-this-go}{%
\subsection{\texorpdfstring{\emph{Literature Survey\ldots{} WHERE DOES
THIS
GO???}}{Literature Survey\ldots{} WHERE DOES THIS GO???}}\label{literature-survey-where-does-this-go}}

To determine how Pagel's \(\lambda\) is commonly utilized in empirical
studies, we conducted a literature survey. From Google.scholar we
obtained a list of all papers published in 2019 that used \(\lambda\);
resulting in 341 studies. For each study, we extracted all
\(\lambda_{est}\), the size of the phylogeny (\(n\)), and noted whether
authors reported confidence intervals or performed significance tests
assessing difference of \(\lambda_{est}\) from either zero or one. We
also noted whether biological interpretations based on \(\lambda_{est}\)
were made, and for studies that reported more than one
\(\lambda_{est}\), we also noted whether these were compared in some
manner, and whether such comparisons were accompanied with statistical
tests between \(\lambda_{est}\).

\emph{Results}. We found 182 manuscripts from 2019 that estimated and
reported Pagel's lambda values using PGLS methods. These papers averaged
8.527 lambda values, ranging from a single lambda estimate up to 71
estimated lambdas. Almost exactly half of the published lambda estimates
were either below 0.05 (25.32\%) or above 0.9 (24.74\%; Figure 3).
73.32\% of the published lambdas were estimated using phylogenies with
fewer than 200 tips, and 348 lambda estimates (8.57\% of all published
estimates) came from phylogenies with fewer than 30 tips. \hfill\break

{[}insert Figure 4 here{]} \hfill\break

Many of the reviewed manuscripts liberally interpreted the magnitude of
the estimate lambda, using phrases such as ``strong'' or ``weak''
phylogentic signal when statistically, all that was clear was a
difference between the estimated lambda and 0 or 1 respectively. We
estimated that about 20.49\% of the manuscripts revealed some sort of
biological interpretation of the magnitude of estimated phylogenetic
signal that overreached the statistical findings. We also identified
seven manuscripts as having inappropriately interpreted differences in
lambda values, indicating that some traits had stronger or weaker signal
than other traits without the appropriate statistical tests.
\hfill\break

As is evidenced by macroevolutionary papers published in 2019 papers,
Pagel's lambda estimation methods are often misused and
over-interpretted. Despite the urging of Boettiger and colleagues to
publish confidence intervals with all lambda parameter estimates, only
18\% of papers published in 2019 do so.

\hypertarget{discussion}{%
\section{Discussion}\label{discussion}}

@@

\textbf{Figure 1}. Precision of Pagel's \(\lambda\) across known levels
of input phylogenetic signal (\(\lambda_{in}\)) on phylogenies of
various sizes. As phylogenies increase in size, variation in
\(\lambda_{in}\) decreases; however the precision is not constant across
the range of input levels (\(\lambda_{in}: 0 \to 1\)), and is highest at
intermediate levels of phylogenetic signal. \hfill\break

\textbf{Figure 2}. Precision of Pagel's \(\lambda\) when incorporated in
phylogenetic regression (\(Y\sim X\)), across known levels of input
phylogenetic signal (\(\lambda_{in}\)) on phylogenies of various sizes.
As phylogenies increase in size, variation in \(\lambda_{in}\)
decreases; however the precision is not constant across the range of
input levels (\(\lambda_{in}: 0 \to 1\)), and is highest at intermediate
levels of phylogenetic signal. \hfill\break

\textbf{Figure 3}. Variation in estimates of phylogenetic signal across
input levels of phylogenetic signal. (A) Estimates of Pagel's
\(\lambda\) for data simulated on phylogenies with 128 taxa (\(n=128\)),
(B) Estimates of \(Z_K\) for data simulated on phylogenies with 128 taxa
(\(n=128\)), (C) Variance in the variation of \(\lambda_{est}\) across
input levels of phylogenetic signal, estimated on phylogenies containing
differing numbers of species. (D) Variance in the variation of \(Z_K\)
across input levels of phylogenetic signal, estimated on phylogenies
containing differing numbers of species. \hfill\break

\textbf{Figure 4}. Frequency of estimated lambda values published in
manuscripts in 2019. The majority of these values were close to 0 or 1,
and from phylogenies with fewer than 200 taxa.

@@

\singlespacing \textbf{Figure 1}. Precision of Pagel's \(\lambda\)
across known levels of input phylogenetic signal (\(\lambda_{in}\)) on
phylogenies of various sizes. As phylogenies increase in size, variation
in \(\lambda_{in}\) decreases; however the precision is not constant
across the range of input levels (\(\lambda_{in}: 0 \to 1\)), and is
highest at intermediate levels of phylogenetic signal.

@@

\singlespacing \textbf{Figure 2}. Precision of Pagel's \(\lambda\) when
incorporated in phylogenetic regression (\(Y\sim X\)), across known
levels of input phylogenetic signal (\(\lambda_{in}\)) on phylogenies of
various sizes. As phylogenies increase in size, variation in
\(\lambda_{in}\) decreases; however the precision is not constant across
the range of input levels (\(\lambda_{in}: 0 \to 1\)), and is highest at
intermediate levels of phylogenetic signal.

@@

\singlespacing \textbf{Figure 3}. Variation in estimates of phylogenetic
signal across input levels of phylogenetic signal. (A) Estimates of
Pagel's \(\lambda\) for data simulated on phylogenies with 128 taxa
(\(n=128\)), (B) Estimates of \(Z_K\) for data simulated on phylogenies
with 128 taxa (\(n=128\)), (C) Variance in the variation of
\(\lambda_{est}\) across input levels of phylogenetic signal, estimated
on phylogenies containing differing numbers of species. (D) Variance in
the variation of \(Z_K\) across input levels of phylogenetic signal,
estimated on phylogenies containing differing numbers of species.

@@

\singlespacing \textbf{Figure 4}.Frequency of estimated lambda values
published in manuscripts in 2019. The majority of these values were
close to 0 or 1, and from phylogenies with fewer than 200 taxa.

>>>>

Evolutionary biologists commonly seek to describe the relative strength
of phylogenetic signal in phenotypic traits, to determine the extent to
which shared evolutionary history has influenced trait covariation among
taxa. This is often accomplished by interpreting empirical estimates of
\(\lambda\); with smaller values signifying `weak' phylogenetic signal,
while larger values are interpreted as `strong' phylogenetic signal
(e.g., De Meester et al. 2019; Pintanel et al. 2019; Su et al. 2019).
Other approaches for interpreting \(\lambda\) are more statistical. For
instance, some have evaluated whether the observed \(\lambda\) differs
from some expected value through the use of confidence intervals
(Vandelook et al. 2019) or by performing likelihood ratio tests that
compare the observed model fit to that obtained when \(\lambda=0\) or
\(\lambda=1\) (Freckleton et al. 2002; Cooper et al. 2010; Bose et al.
2019). Additionally, qualitative comparisons of \(\lambda\) estimates
obtained from multiple phenotypic traits have been used to infer whether
the strength of phylogenetic signal is greater in one trait as compared
to another (e.g., Liu et al. 2019; Bai et al. 2019). Indeed, statements
regarding the strength of phylogenetic signal based on \(\lambda\) are
rather common in the evolutionary literature. We conducted a literature
survey in Google.scholar and found that of the 204 papers published in
2019 that estimated and reported Pagel's \(\lambda\), 40\% interpreted
the strength of phylogenetic signal for at least one phenotypic trait.
Additionally, nearly 30\% of the 421 \(\lambda\) estimates between 0.25
and 0.75 were assigned a strength of signal, and because nearly half of
the 1,572 values reported were near the limits of the parameter (Figure
1), this percentage is even higher, as biological interpretation of
phylogenetic signal at the limits of \(\lambda\) are known. \hfill\break

It seems intuitive to interpret the strength of phylogenetic signal
based on the value of \(\lambda\), as \(\lambda\) is a parameter on a
bounded scale (\(0\to1\)) for which interpretation of its extremal
points are understood. Specifically, \(\lambda=0\) represents no
phylogenetic signal, while \(\lambda=1\) is phylogenetic signal as
expected under Brownian motion. However, equating values of \(\lambda\)
directly to the strength of phylogenetic signal presumes two important
statistical properties that have not been fully explored. First, it
presumes that values of \(\lambda\) can be precisely estimated, as
biological inferences regarding the strength of phylogenetic signal
depend on high accuracy in its estimation. Therefore, understanding the
precision in estimating \(\lambda\) is paramount. One study (Boettiger
et al. 2012) found that estimates of Pagel's \(\lambda\) displayed less
variation (i.e., greater precision) when data were simulated on a large
phylogeny (\(N=281\)) as compared to a small one (\(N=13\)). From this
observation it was concluded that insufficient data (i.e., the number of
species) was the underlying cause of the increased variation across
parameter estimates (Boettiger et al. 2012). Indeed, such a pattern is
common with statistical estimators, as summary statistics and parameters
are often more precise at greater sample sizes (Cohen 1988). However,
this conclusion also assumes that the precision of \(\lambda\) remains
constant across its range (\(\lambda = 0 \to 1\)); an assumption that to
date, has not been verified. Thus, despite widespread use of Pagel's
(1999) \(\lambda\) in macroevolutionary studies, at present, we still
lack a general understanding of the precision with which \(\lambda\) can
estimate levels of phylogenetic signal in phenotypic datasets.
\hfill\break

@@

\emph{Results.} We found that the precision of \(\lambda_{est}\) varied
widely across simulation conditions. Predictably, precision improved as
the number of species increased (Figure 2). This confirmed earlier
findings of Boettiger et al.~(2012), and adhered to parametric
statistical theory. However, in many cases the set of \(\lambda_{est}\)
spanned nearly the entire range of possible values (e.g., \(n=32\);
\(\lambda_{in}=0.5\): \(\lambda_{est}= 0.0\to 0.985\)), revealing that
estimates of \(\lambda\) were not a reliable indicator of input
phylogenetic signal. Importantly, the precision of \(\lambda_{est}\) was
not uniform across all levels of phylogenetic signal, with the worst
precision at intermediate levels of signal (\(\lambda_{in}\approx0.5\)),
and improved precision as input levels approached the extremes of its
range (i.e., \(\lambda_{in}\rightarrow0\) \&
\(\lambda_{in}\rightarrow1\)). Thus, estimates of \(\lambda\) were least
reflective of the true input signal at intermediate values.
Additionally, even at large levels of species richness, we found that
the range of \(\lambda_{est}\) still encompassed a substantial portion
of possible values (e.g., \(n=512\); \(\lambda_{in}=0.5\):
\(\lambda_{est} = 0.32\to 0.68\)). Likewise, the same \(\lambda_{est}\)
could be obtained from datasets containing vastly different input levels
of phylogenetic signal (e.g., \(n=512\); \(\lambda_{est} = 0.5\);
\(\lambda_{in} = 0.25 \to 0.65\)). Results were similar when \(\lambda\)
was co-estimated with regression parameters in PGLS regression (Figure
3). Here, regression parameters (\(\beta\)) were accurately estimated,
confirming earlier findings of Revell 2010 (2010) (see Supporting
Information). However, estimates of phylogenetic signal were not, and
the spread of \(\lambda_{est}\) was even broader than that observed when
\(\lambda\) was estimated for only the dependent variable. Taken
together, these findings reveal that \(\lambda_{est}\) does not
precisely characterize observed levels of phylogenetic signal in
phenotypic datasets, and that biological interpretations of the strength
of phylogenetic signal based on \(\lambda\) may be highly inaccurate.
\hfill\break

@@

{[}insert Figure 3 here{]} \hfill\break

@@

Adams and Collyer (2019b) recently suggested that the strength of
phylogenetic signal could be represented as an effect size, based on the
\(Kappa\) statistic and its empirical sampling distribution from
permutation. Here we formalize that suggestion, and find an effect size
as:

@@

where \(K_{obs}\) is the observed phylogenetic signal, and \(\hat\mu_K\)
and \(\hat\sigma_K\) are the mean and standard deviation of the
empirical sampling distribution of \(Kappa\) obtained via permutation.
Similarly, an effect size based on \(\lambda\) could be envisioned as:

\begin{align}
    Z_{\lambda}=\frac{\lambda_{obs}-0}{\hat\sigma_\lambda}.
\end{align}

In this case, \(\lambda_{obs}\) and \(\hat\sigma_\lambda\) are
empirically derived using maximum likelihood. Note also that under the
null hypothesis, \(E(\lambda)=0\), a no phylogenetic signal is expected
under this condition (Freckleton et al. 2002). \hfill\break

To evaluate the utility of \(Z_K\) and \(Z_\lambda\) we calculated both
effect sizes for the simulated datasets generated above, and summarized
the precision of each using its variance (\(\sigma^2_{Z_K}\) and
\(\sigma^2_{Z_\lambda}\)). Results are found in Figure 4. Here two
things are evident. First, estimates of \(Z_K\) track the input
phylogenetic signal in a more linear fashion than do estimates of
\(Z_\lambda\). Second, the precision of \(Z_K\) is considerably more
stable as compared with \(Z_\lambda\), as coefficients of variation for
the set of \(\sigma^2_{Z_K}\) across input levels of phylogenetic signal
were an order of magnitude smaller for than was observed for
\(\sigma^2_{Z_\lambda}\) (Figure 4). This implied that estimates of the
strength of phylogenetic signal were more reliable and robust when using
\(Z_K\) as compared with \(Z_\lambda\). \hfill\break

{[}insert Figure 4 here{]}

@@

@@

\textbf{Figure 1}. Frequency distribution of \(\lambda\) estimates
published in 2019. The majority of these values were close to 0 or 1,
and from phylogenies with fewer than 200 taxa. \hfill\break

\textbf{Figure 2}. Precision of Pagel's \(\lambda\) across known levels
of input phylogenetic signal (\(\lambda_{in}\)) on phylogenies of
various sizes. As phylogenies increase in size, variation in
\(\lambda_{in}\) decreases; however the precision is not constant across
the range of input levels (\(\lambda_{in}: 0 \to 1\)), and is highest at
intermediate levels of phylogenetic signal. \hfill\break

\textbf{Figure 3}. Precision of Pagel's \(\lambda\) when incorporated in
phylogenetic regression (\(Y\sim X\)), across known levels of input
phylogenetic signal (\(\lambda_{in}\)) on phylogenies of various sizes.
As phylogenies increase in size, variation in \(\lambda_{in}\)
decreases; however the precision is not constant across the range of
input levels (\(\lambda_{in}: 0 \to 1\)), and is highest at intermediate
levels of phylogenetic signal. \hfill\break

\textbf{Figure 4}. Variation in estimates of phylogenetic signal across
input levels of phylogenetic signal. (A) Estimates of Pagel's
\(\lambda\) for data simulated on phylogenies with 128 taxa (\(n=128\)),
(B) Estimates of \(Z_K\) for data simulated on phylogenies with 128 taxa
(\(n=128\)), (C) Variance in the variation of \(\lambda_{est}\) across
input levels of phylogenetic signal, estimated on phylogenies containing
differing numbers of species. (D) Variance in the variation of \(Z_K\)
across input levels of phylogenetic signal, estimated on phylogenies
containing differing numbers of species.

@@

\singlespacing \textbf{Figure 1}. Frequency distribution of \(\lambda\)
estimates published in 2019. The majority of these values were close to
0 or 1, and from phylogenies with fewer than 200 taxa.

@@

\singlespacing \textbf{Figure 2}. Precision of Pagel's \(\lambda\)
across known levels of input phylogenetic signal (\(\lambda_{in}\)) on
phylogenies of various sizes. As phylogenies increase in size, variation
in \(\lambda_{in}\) decreases; however the precision is not constant
across the range of input levels (\(\lambda_{in}: 0 \to 1\)), and is
highest at intermediate levels of phylogenetic signal.

@@

\singlespacing \textbf{Figure 3}. Precision of Pagel's \(\lambda\) when
incorporated in phylogenetic regression (\(Y\sim X\)), across known
levels of input phylogenetic signal (\(\lambda_{in}\)) on phylogenies of
various sizes. As phylogenies increase in size, variation in
\(\lambda_{in}\) decreases; however the precision is not constant across
the range of input levels (\(\lambda_{in}: 0 \to 1\)), and is highest at
intermediate levels of phylogenetic signal.

@@

\singlespacing \textbf{Figure 4}. Variation in estimates of phylogenetic
signal across input levels of phylogenetic signal. (A) Estimates of
Pagel's \(\lambda\) for data simulated on phylogenies with 128 taxa
(\(n=128\)), (B) Estimates of \(Z_K\) for data simulated on phylogenies
with 128 taxa (\(n=128\)), (C) Variance in the variation of
\(\lambda_{est}\) across input levels of phylogenetic signal, estimated
on phylogenies containing differing numbers of species. (D) Variance in
the variation of \(Z_K\) across input levels of phylogenetic signal,
estimated on phylogenies containing differing numbers of species.
